An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-07-22 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

\newpage
# Introduction
The bacterial family *Erwiniaceae* includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, and *Kalamiella* [REF CRAIG]. *Mixta* is a relatively newer genus first identified by Palmer et al. in 2017 [@Palmer2017] and later described by the same researchers in 2018 [@Palmer2018]. Currently, there are only five *Mixta* species validly published [@LPSN]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. The other *Mixta* species include *M. theicola* which was isolated from black tea extract [@Kato2015], *M. intestinalis* which was isolated from a healthy human fecal sample [@Prakash2015], and *M. tenebrionis* which was isolated from the gut microbiota of a plastic-eating mealworm [@Xia2020]. The first two were previously named as *Pantoea* species.

*Mixta*'s position relative to other members of the *Erwiniaceae*, in particular to *Pantoea*, *Erwinia*, *Tatumella*, differs vastly between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used *gyrB*, *rpoB*, *atpD*, and *infB* nucleotide gene sequences in their phylogenetic analyses. Prakash et al. [@Prakash2015] conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method [@Palmer2017; @Rezzonico2016; @Gueule2015]. Three of the studies – Rezzonico et al. [@Rezzonico2016], Prakash et al. [@Prakash2015], and Gueule et al. [@Gueule2015] – placed the *Mixta* species within the *Pantoea* genera. In contrast, the analysis in Palmer et al. [@Palmer2017] identified *Mixta* outside of the other three genera. A separate study by Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. *Mixta* was found to be related closest to *Tatumella* by Rezzonico et al. [@Rezzonico2016] and *Pantoea* by Palmer et al. [@Palmer2018]. Brady et al. [@Brady2012] suggested that *Mixta* was a part of the *Pantoea* genus and not separate. Additionally, Palmer et al. [@Palmer2017] computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*.

@@

## A Comparison of the Four Analyses
The four above analyses (genetic distance and site identity, both using NTS and AAS) gave slightly different results for the *Mixta* sequences. The four analyses agreed upon the closest relative for approximately 38.7% of each of the *M. calida* and *M. gaviniae* sequences. The number of sequences closely related to one of the other taxa differed between the two *Mixta* species (Table 3). Additionally, although 38.7% of both *Mixta* species’ sequences had a consistent closest relative, 30 genes differed between them. For these 30 genes, there was a consensus for only one of the *Mixta* species and not the other. Nevertheless, 34.9% of the *M. calida* and *M. gaviniae* homologues agreed on the same species.

@@

## Phylogenetic Trees Showing Other Results
Several genes were found to be more closely related to taxa other than *P. septica* or *P. agglomerans*. Thus, gene sequences wherein the four analyses agreed upon each of the other taxa were concatenated to compute phylogenetic trees. There was a consensus for 32 genes that the closest relative was not either of the *Pantoea* species; 17 were more closely related to *E. tasmaniensis*, 8 to *E. amylovora*, 1 to *T. saanichensis*, and 6 to *E. cloacae*.

@@

Phylogenetic analysis is an accepted method for understanding the evolutionary relatedness [@Boundless2021] while site identity is used to confirm identification and taxon boundaries [@Ciufo2018]. These two methods were conducted in this study as both nucleotide (NTS) and amino acid (AAS) sequences. *P. septica* was the closest relative to the majority of *Mixta* homologues in all four analyses. This lends strong support that *Mixta* and *P. septica* share a recent common ancestor, followed by other *Pantoea* species, *Erwinia*, and then *Tatumella*. This agrees with *Mixta* being previously known as part *Pantoea*. Additionally, there were no *Mixta* sequences that were closely related to *P. syringae* which demonstrates that this species is a suitable outgroup.

For both methods, fewer genes were labelled as being closely related to *P. septica* when using the AAS versus the NTS. This difference may be to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy. Nevertheless, sequence type, as well as different evolutionary models between NTS and AAS, affects the results. AAS may be better suited for uncovering taxonomic relationships at the genus or higher level and NTS may be more appropriate for the species level.

When considering the spatial effects of the closest relatives, the uniform pattern of the genes related to *P. septica* further lends support that this species and *Mixta* share a recent common ancestor. Additionally, the other taxa followed the same pattern. If the genes that were closely related to non-*Pantoea* species are a result of HGT, then this suggests that HGT may be a common occurrence within the *Erwiniaceae*. A second explanation for the even distribution patterns in the closest relatives is that the genes are highly conserved. Two or more species may have been tied for closest relative and the R program may have selected one based another characteristic, such as alphabetical order. Nevertheless, further research into these genes and the frequency of HGT events with *Mixta* and other genera within the *Erwiniaceae* is needed.

# Contributions to the Field Statement

>>>>

# Introduction
The bacterial family *Erwiniaceae* includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, and *Kalamiella* [REF CRAIG]. *Mixta* is a relatively newer genus first identified by Palmer et al. in 2017 [@Palmer2017] and later described by the same researchers in 2018 [@Palmer2018]. Currently, there are only five *Mixta* species validly published [@LPSN]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. The other *Mixta* species include *M. theicola* which was isolated from black tea extract [@Kato2015], *M. intestinalis* which was isolated from a healthy human fecal sample [@Prakash2015], and *M. tenebrionis* which was isolated from the gut microbiota of a plastic-eating mealworm [@Xia2020]. The first two were previously named as *Pantoea* species.

*Mixta*'s position relative to other members of the *Erwiniaceae*, in particular to *Pantoea*, *Erwinia*, *Tatumella*, differs vastly between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used *gyrB*, *rpoB*, *atpD*, and *infB* nucleotide gene sequences in their phylogenetic analyses. Prakash et al. [@Prakash2015] conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method [@Palmer2017; @Rezzonico2016; @Gueule2015]. Three of the studies – Rezzonico et al. [@Rezzonico2016], Prakash et al. [@Prakash2015], and Gueule et al. [@Gueule2015] – placed the *Mixta* species within the *Pantoea* genus. In contrast, the analysis in Palmer et al. [@Palmer2017] identified *Mixta* outside of the other three genera. A separate study by Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA (multi-locus sequence analysis) genes as amino acid sequences. *Mixta* was found to be related closest to *Pantoea* by Palmer et al. [@Palmer2018]. Both Rezzonico et al. [@Rezzonico2016] and Brady et al. [@Brady2012] suggested that *Mixta* was a part of the *Pantoea* genus and not separate. Additionally, Palmer et al. [@Palmer2017] computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*.

@@

## A Comparison of the Four Analyses
The four above analyses (genetic distance and site identity, both using NTS and AAS) gave slightly different results for the *Mixta* sequences. The four analyses agreed upon the closest relative for approximately 38.7% of each of the *M. calida* and *M. gaviniae* sequences. The number of sequences closely related to one of the other taxa differed between the two *Mixta* species (Table 3). Additionally, although these sequences had a consistent closest relative, 30 genes differed between them. For these 30 genes, there was a consensus for only one of the *Mixta* species and not the other. Nevertheless, 34.9% of the *M. calida* and *M. gaviniae* homologues agreed on the same species.

@@

## Phylogenetic Trees Showing Other Results
Several genes were found to be more closely related to taxa other than *P. septica* or *P. agglomerans*. Thus, gene sequences wherein the four analyses agreed upon each of the other taxa were concatenated to compute phylogenetic trees. There was a consensus for 32 genes that the closest relative was not either of the *Pantoea* species: 17 were more closely related to *E. tasmaniensis*, 8 to *E. amylovora*, 1 to *T. saanichensis*, and 6 to *E. cloacae*.

@@

Phylogenetic analysis is an accepted method for understanding the evolutionary relatedness [@Boundless2021], while site identity is used to confirm identification and taxon boundaries [@Ciufo2018]. These two methods were conducted in this study as both nucleotide (NTS) and amino acid (AAS) sequences. *P. septica* was the closest relative to the majority of *Mixta* homologues in all four analyses. This lends strong support that *Mixta* and *P. septica* share a recent common ancestor, followed by other *Pantoea* species, *Erwinia*, and then *Tatumella*. This result agrees with *Mixta* being previously known as part *Pantoea*. Additionally, there were no *Mixta* sequences that were closely related to *P. syringae*, which demonstrates that this species is a suitable outgroup.

For both methods, fewer genes were labelled as being closely related to *P. septica* when using the AAS versus the NTS. This difference may be to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy. Nevertheless, sequence type and different evolutionary models between NTS and AAS affect the results. AAS may be better suited for uncovering taxonomic relationships at the genus or higher level, and NTS may be more appropriate for the species level.

When considering the spatial effects of the closest relatives, the uniform pattern of the genes related to *P. septica* further lends support that this species and *Mixta* share a recent common ancestor. Additionally, the other taxa followed the same pattern. If the genes closely related to non-*Pantoea* species are a result of HGT, this suggests that HGT may be a common occurrence within the *Erwiniaceae*. A second explanation for the consistent distribution patterns in the closest relatives is that the genes are highly conserved. Two or more species may have been tied for the closest relative, and the R program may have selected one based on another characteristic, such as alphabetical order. Nevertheless, further research into these genes and the frequency of HGT events with *Mixta* and other genera within the *Erwiniaceae* is needed. Furthermore, for 279 (34.9%) homologues, the closest relative was consistent across all four analyses and between the two *Mixta* species. These results are more likely to be accurate because there was a strong phylogenetic signal and conserved site identity between the NTS and AAS. An additional 30 genes differed between *M. calida* and *M. gaviniae* on the closest species relative where there was a consensus for only one of the *Mixta* species. Codon degeneracy, codon bias, and HGT to one of the *Mixta* species may explain contentious results between the analyses and between the *Mixta* species.

The closest genus relative was also considered to account for sequence conservation and similar identity in recently diverged species. There was an agreement on the closest genus relative across the analyses for approximately 63% of the *Mixta* sequences. Of these, only 39 homologues differed between the *Mixta* species, with only one gene, *ttuB*, having a consensus on the closest genus relative for both of the *Mixta* species. A gene tree for *ttuB* reveals a possible HGT event wherein *M. calida* received a homologue from a *Tatumella* species. Because *ttuB* is a single-copy gene, the original version may have been deleted. Additionally, this explains why the genetic distance results for *M. calida* and *M. gaviniae* differed for both the NTS and AAS. A second gene, *secY*, also gave contentious results between the *Mixta* species in the AAS genetic distance analysis; however, this is likely due to its highly conservative nature given the short branch lengths in its gene tree.

Many previous studies into the *Erwiniaceae* have conducted phylogenetic analysis using four core genes: *atpD*, *gyrB*, *infB*, and *rpoB* [@Prakash2015; @Palmer2017; @Rezzonico2016; @Gueule2015]. Two genes, *atpD* and *gyrB*, were removed from this study’s dataset during the filtering process; however, *infB*, *rpoB*, and twelve other related subunits were available. The concatenated NTS sequences produced a similar topology tree to Brady et al.’s [@Brady2011] maximum-likelihood (ML) tree based on 16S rRNA gene sequences, and none of the trees based on the four genes. The differences may be caused by different software programs, phylogenetic models, and the addition of genes in my study. Additionally, the topology of the AAS sequences of the 14 core genes was not found by any previous studies [@Prakash2015; @Palmer2017; @Palmer2018; @Rezzonico2016; @Gueule2015; @Brady2011; @Brady2012]. When the four genes were concatenated as AAS, Palmer et al. [@Palmer2018] found *Pantoea* and *Tatumella* formed sister taxa followed by *Mixta*. My analysis identified *Pantoea* and *Erwinia* as sister taxa followed by *Mixta*. Nonetheless, the lack of branch support leaves the true relationships ambiguous for the AAS.

Interestingly, phylogenetic trees showing a close relationship between *Mixta* and a non-*Pantoea* species were highly supported for both the NTS and AAS. Genes wherein there was a consensus between the four analyses were concatenated to create phylogenetic trees for each of *E. tasmaniensis*, *E. amylovora*, *T. saanichensis*, and *E. cloacae*. The strong branch support and long branch lengths suggest that these genes are likely candidates for HGT. Furthermore, the inclusion of these genes into future phylogenetic studies of the *Erwiniaceae* may result in contentious results and add a phylogenetic signal that is not due to vertical evolutionary history.

In conclusion, in phylogenetic analyses of the *Erwiniaceae*, diligence must be given to the software program, evolutionary model, sequence type, and included genes. Different programs may offer other methods, models, and parameters. Additionally, model testing is a valuable tool to ensure the most accurate estimation of evolutionary history is calculated. The sequence type – nucleotides versus amino acids – further play a role due to the difference in sequence lengths and codon degeneracy. The selected genes may also affect phylogenetic results as conserved genes may not conclusively separate species, and HGT genes may skew the taxa’s relationships. Traditional phylogenetic methods (i.e., genetic distance and phylogenetic trees) versus site identity may also give contentious results. Phylogenetic analyses estimate evolutionary history, and some models penalise specific mutations. However, site identity analyses provide the similarity between sequences and operate on the assumption that members of a clade should be genetically similar. Ultimately, it is best practice for future studies to be transparent in how they conducted their phylogenetic analyses to ensure reproducibility.

<!-- # Author Contributions (please edit if I accidentally missed something) -->
<!-- KH carried out conceptualisation, methodology, investigation, formal analysis, and writing (original draft preparation). JS carried out conceptualisation, methodology, resources (study materials and computing resources), writing (review and editing), and supervision. -->
